% !TEX root = ../PhD Thesis.tex

\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

My PhD has been an incredible journey where I met fantastic people, some of which I could never imagine living without. It was a journey where I faced difficult challenges, some of which made me the person I am today. But no matter the obstacles, I am glad that I walked this path full of mesmerising moments and immeasurable learning.

First, a big thank you to Nuno Morais for teaching me during all these years and for inviting me to his lab and iMM, opening the doors to this crazy adventure that now comes to an end. I also want to thank my former lab mates for all the scientific discussions and scientifically-proven fun, specially Marie and Lina that are inspirational people with big smiles and even bigger hearts.

iMM is a remarkable research institute with a vibrant community composed of minds that are as brilliant as they are supportive. Thank you to all those friends that I intend to continue hanging out and sharing my life with: Inês Mendes, Elvira, Eunice, Helena, Filipe, João, Madalena, Inês Faleiro and Ana Duarte. For all the nights out, partying out throughout my PhD and teaching me invaluable lessons while holding one or more beers, thank you Debanjan and Vasco.

To my greatest chaperon over this whole journey, thank you Mariana. No matter how hard the journey was, I could always count on you to share any thoughts, frustrations, achievements, books, computer fonts, colours and nerdy facts. I can only thank you for our time senescing together.

I will carry all my friends in my heart, including those that were already there for a long time: Alicia and João. Our friendship knows no distance. No matter where each one of us lives, we will always be together.

Last, but not least, one special thank you to my mother, father, sister, brother and my whole family. To my grandfather, my biggest supporter even before I was born and the person that made me love biology and science, I dedicate my PhD thesis -- I would not be here today if it was not for you. Thank you.

It took millions and millions of years for the Earth to cool down, for the first life forms to reproduce and for mammals to evolve so I could be where I am today. No wonder I feel like these 4 years went by in the blink of a supernova. Although unsure of what lies ahead, I know that I can count on my friends and family to cheer me up. To many more years of friendship, learning and living! See you all in my next step.